---
title: "NOAA Storm Data Analysis - Most Harmful and Most Expensive Storm Events"
author: "Eddo W. Hintoso"
date: "August 14, 2015"
output: 
  html_document:
    keep_md: true
---

# Synopsis
This data analysis attempts to address the following questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?

The U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database was analysed to answer these questions. Storm data from 1996 to 2011 was used, in order to minimize bias due to limited data being collect in previous years. Using recent data also minimized the effects of inflation on the cost of the damage due to the events.

To see the software conditions used to complete this data analysis, see the section **Software Conditions**.

**NOTE**: The global options are set to cache the computations so that the `.Rmd` file can be knitted to HTML faster.
```{r setoptions}
library(knitr)
opts_chunk$set(cache = TRUE)
```

# Data Processing

## Validating Pre-Existence of Data
```{r}
##  check if file exists; if not, download the file again
if(!file.exists("repdata-data-StormData.csv.bz2")){
    message("Downloading data")
    fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(fileURL, destfile = "./repdata-data-StormData.csv.bz2", method = "curl")
}
```

## Reading the Data
```{r}
##  read data
storm <- read.csv('repdata-data-StormData.csv.bz2')

##  store it as 'tbl_df' for faster processing
library(dplyr)
storm <- tbl_df(storm)
```

## Cleaning the Data
According to the [information](http://www.ncdc.noaa.gov/stormevents/details.jsp) on the Storm Events Database provided, the events types available are:  
1. *Tornado*: From 1950 through 1954, only tornado events were recorded.  
2. *Tornado, Thunderstorm Wind and Hail*: From 1955 through 1992, only tornado, thunderstorm wind and hail events were keyed from the paper publications into digital data. From 1993 to 1995, only tornado, thunderstorm wind and hail events have been extracted from the [Unformatted Text Files](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=collection).  
3. *All Event Types (48 from Directive 10-1605)*: From 1996 to present, 48 event types are recorded as defined in [NWS Directive 10-1605](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf). 

As mentioned in the synopsis before, only storm data from 1996 to 2011 was used, in order to minimize bias due to limited data being collect in previous years. Using recent data also minimized the effects of inflation on the cost of the damage due to the events. But first, we need to check if the dates stored are in the `Date` format.
```{r}
##  transform BGN_DATE into Date format
if (class(storm$BGN_DATE) != "Date") {
    storm <- tbl_df(transform(storm, BGN_DATE = as.Date(storm$BGN_DATE, format = "%m/%d/%Y")))
}

##  confirm BGN_DATE is in Date format
class(storm$BGN_DATE)
```

Now, to filter out the data before 1996,
```{r}
##  set cutoff date and filter
dateCutoff <- as.Date("1996-01-01", format = "%Y-%m-%d")
storm <- subset(storm, BGN_DATE >= dateCutoff)

##  confirm date is from 1996-01-01 onwards
head(storm$BGN_DATE, 4)
```
Since the dates are listed in chronological order, we can see that the data is indeed only 1996 onwards.

### Column Filter
Needless to say, some variables in the data would not contribute to the analysis that would answer the questions posed. Therefore, those columns have to be filtered out as well - but not before checking every variable and evaluating which ones to keep.
```{r}
##  display all variables
names(storm)
```
From this, we can decide to keep only the variables shown below and accordingly filter into a new variable `newStorm`:
```{r}
##  create character vector of useful variables
neededCols <- c("BGN_DATE", "STATE", "EVTYPE", "BGN_LOCATI", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "REMARKS", "REFNUM")

##  filter variables in storm data
newStorm <- storm[, neededCols]

##  confirm variable column filter
names(newStorm)
```

### Row Filter
The numeric values of columns `FATALITIES`,`INJURIES`,`PROPDMG`, and `CROPDMG`, will be no of use when all of them are equal to 0, since it is quantifiably speaking, not contributing to the harm of population health or the economy. That being said, certain rows will be filtered out from the data.
```{r}
##  remove rows with no damage whatsoever
newStorm <- subset(newStorm, !(FATALITIES<=0 & INJURIES<=0 &PROPDMG<=0 & CROPDMG<=0))
```

### Event Type Regroup
Let us examine the `EVTYPE` variable and see how many different event types we get out of it.
```{r}
##  list every different event types
unique(newStorm$EVTYPE)
```
**`r length(unique(newStorm$EVTYPE))`** different event types sound like the data can be pretty specific with their labels. However, there are many potential regroupings that can be performed. The issues here are that some of them are not all capitalized, and that some have spaces here and there. The following one line of code should remedy the problem:
```{r}
##  transform column variable into uppercase character 
newStorm$EVTYPE <- gsub("^ *", "",toupper(as.character(newStorm$EVTYPE)))
```
Then, some event types are abbreviated (or over-specified) when in fact it can potentially be regrouped as one, i.e. `TSTM WIND` is presumably indentical to `THUNDERSTORM WIND`. For convenience's sake, I plan to group any event type containing `TSTM` and `THUNDERSTORM` as simply `THUNDERSTORM`.
```{r}
##  regoup as one
TSTM <- grep("TSTM", newStorm$EVTYPE)
THUNDERSTORM <- grep("THUNDERSTORM", newStorm$EVTYPE)
newStorm$EVTYPE[TSTM] <- "THUNDERSTORM"
newStorm$EVTYPE[THUNDERSTORM] <- "THUNDERSTORM"

##  confirm shortened event type list
unique(newStorm$EVTYPE)
```
As one can see, the number of event types were reduced from **222** from **`r length(unique(newStorm$EVTYPE))`**.  
Now, by inspection, other potential regroups involve FLD being abbreviated from FLOOD, and COLD, HEAT, DRY, and SNOW having labels that are overly specific. Needless to say, categorization can be done in many ways, but the demonstrated method is sufficiently simple, in order that redundacy and over-specifity is minimized. The following changes are made (with personal discretion) to regroup the rest of the list:
```{r}
##  identify key words
FLD <- grep("FLD", newStorm$EVTYPE)
FLOOD <- grep("FLOOD", newStorm$EVTYPE)
COLD <- grep("COLD", newStorm$EVTYPE)
DRY <- grep("DRY", newStorm$EVTYPE)
HEAT <- grep("HEAT", newStorm$EVTYPE)
SNOW <- grep("SNOW", newStorm$EVTYPE)
RAIN <- grep("RAIN", newStorm$EVTYPE)
RIP_CURRENT <- grep("RIP CURRENT", newStorm$EVTYPE)

##  regroup accordingly
newStorm$EVTYPE[FLD] <- "FLOOD"
newStorm$EVTYPE[FLOOD] <- "FLOOD"
newStorm$EVTYPE[COLD] <- "COLD"
newStorm$EVTYPE[DRY] <- "DRY"
newStorm$EVTYPE[HEAT] <- "HEAT"
newStorm$EVTYPE[SNOW] <- "SNOW"
newStorm$EVTYPE[RAIN] <- "RAIN"
newStorm$EVTYPE[RIP_CURRENT] <- "RIP CURRENT"

##  examine current list of event types
unique(newStorm$EVTYPE)
```
The list seems pretty shortened right now, only approximately **`r as.integer(length(unique(newStorm$EVTYPE)) / 222 * 100)`**% of its original length.

Of course, one can argue to regroup Wind, which can be defined and labeled pretty broadly. Let's examine whether Wind would be a good keyword for regrouping.
```{r}
##  search Wind as keyterm through list of event types (use unique function)
wind <- grep("WIND", unique(newStorm$EVTYPE))
unique(newStorm$EVTYPE)[wind]
```
As one can see, winds can have different levels of severity. Thus, in this data analysis, it is decided that wind is not regrouped as one. However, keep in mind that the displayed preference of categorization is subjective at best.

# Results

## Types of events that are most harmful with respect to Population Health
The two variables that contribute to population health are `FATALITIES` and `INJURIES`.
```{r}
##  sum all the data up and save new data into new variable
harmHealth <- with(newStorm, aggregate(list(Total_Fatalities = FATALITIES,
                                            Percent_Fatality = 0,
                                            Total_Injuries = INJURIES,
                                            Percent_Injury = 0),
                                       list(Event_Type = EVTYPE),
                                       sum))

##  define percentage rates
harmHealth$Percent_Fatality <- with(harmHealth,
                                    round(Total_Fatalities / sum(Total_Fatalities) * 100, digits = 3))
harmHealth$Percent_Injury <- with(harmHealth,
                                    round(Total_Injuries / sum(Total_Injuries) * 100, digits = 3))

##  arrange data in descending order of total fatalities
fatalities <- arrange(harmHealth, desc(Total_Fatalities))
injuries <- arrange(harmHealth, desc(Total_Injuries))

##  display top 10 for inflicting fatality
topFatalities <- head(fatalities, 10)
topFatalities

##  display top 10 for inflicting injuries
topInjuries <- head(injuries, 10)
topInjuries
```

## Types of events that have the greatest economic consequences
There are two types of damage categorise. Damage to property (PROPDMG) and damage to crops (CROPDMG). These are in two separate fields, with multipliers (PROPDMGEXP, CROPDMGEXP) that indicate how much to multiply the damage field by.

There are three values for the multiplier.  
- K = Thousands  
- M = Millions  
- B = Billions  

Calculate the total cost of damage by adding the cost of property damage and crop damage.
```{r}
##  load 'car' package for recode function
library(car)

##  define numeric converter
numConvert <- "'B'=1000000000; 'M'=1000000; 'K'=1000"

##  transform variable columns of multipliers to character form
newStorm$PROPDMGEXP <- as.character(newStorm$PROPDMGEXP)
newStorm$CROPDMGEXP <- as.character(newStorm$CROPDMGEXP)

##  create new variable column for numeric multiplier
newStorm$PROPDMGMULT <- as.numeric(recode(newStorm$PROPDMGEXP, numConvert))
newStorm$CROPDMGMULT <- as.numeric(recode(newStorm$CROPDMGEXP, numConvert))

##  create new variable column for total damage incurred
newStorm$PROPDMGTOT <- with(newStorm, PROPDMG * PROPDMGMULT)
newStorm$CROPDMGTOT <- with(newStorm, CROPDMG * CROPDMGMULT)
newStorm$DMGTOT <- with(newStorm, PROPDMGTOT + CROPDMGTOT)
```

# Software & Hardware Conditions
```{r}
sessionInfo()
```