---
title: "NOAA Storm Data Analysis - Most Harmful and Most Expensive Storm Events"
author: "Eddo W. Hintoso"
date: "August 14, 2015"
output: html_document
    keep_md: true
---

# Synopsis
This data analysis attempts to address the following questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?

The U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database was analysed to answer these questions. Storm data from 1996 to 2011 was used, in order to minimize bias due to limited data being collect in previous years. Using recent data also minimized the effects of inflation on the cost of the damage due to the events.

To see the software conditions used to complete this data analysis, see the section **Software Conditions**.

**NOTE**: The global options are set to cache the computations so that the `.Rmd` file can be knitted to HTML faster.
```{r setoptions}
library(knitr)
opts_chunk$set(cache = TRUE)
```

# Data Processing

## Validating Pre-Existence of Data
```{r}
##  check if file exists; if not, download the file again
if(!file.exists("repdata-data-StormData.csv.bz2")){
    message("Downloading data")
    fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(fileURL, destfile = "./repdata-data-StormData.csv.bz2", method = "curl")
}
```

## Reading the Data
```{r}
##  read data
storm <- read.csv('repdata-data-StormData.csv.bz2')

##  store it as 'tbl_df' for faster processing
library(dplyr)
storm <- tbl_df(storm)
```

## Cleaning the Data
According to the [information](http://www.ncdc.noaa.gov/stormevents/details.jsp) on the Storm Events Database provided, the events types available are:  
1. *Tornado*: From 1950 through 1954, only tornado events were recorded.  
2. *Tornado, Thunderstorm Wind and Hail*: From 1955 through 1992, only tornado, thunderstorm wind and hail events were keyed from the paper publications into digital data. From 1993 to 1995, only tornado, thunderstorm wind and hail events have been extracted from the [Unformatted Text Files](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=collection).  
3. *All Event Types (48 from Directive 10-1605)*: From 1996 to present, 48 event types are recorded as defined in [NWS Directive 10-1605](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf). 

As mentioned in the synopsis before, only storm data from 1996 to 2011 was used, in order to minimize bias due to limited data being collect in previous years. Using recent data also minimized the effects of inflation on the cost of the damage due to the events. But first, we need to check if the dates stored are in the `Date` format.
```{r}
##  transform BGN_DATE into Date format
if (class(storm$BGN_DATE) != "Date") {
    storm <- transform(storm, BGN_DATE = as.Date(storm$BGN_DATE, format = "%m/%d/%Y"))
}

##  confirm BGN_DATE is in Date format
class(storm$BGN_DATE)
```

# Results

# Software & Hardware Conditions
```{r}
sessionInfo()
```